---
title: "PHW251 - F21_midterm_prep"
author: "Avery Richards"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
# load tidyverse
library(tidyverse)

# load penguins dataset
install.packages("palmerpenguins")
library(palmerpenguins)

```

Part 1. Import and explore the data. Use load(penguins) to load the relevant file, and change island, species, and sex to character (they default to factor)

```{r import_summary}
# make a seperate object of the penguins data, change factors to character format, like stringsAsFactors = FALSE
penguins_df <- penguins %>%  
  mutate(island = as.character(island),
         species = as.character(species),
         sex = as.character(sex)) 

```

Return a summary of the distribution of penguin body mass observations

```{r}

summary(penguins_df$body_mass_g)

```

Return the unique names of species of penguins observed in the study
```{r}

penguins_df %>% 
  distinct(species)

unique(penguins_df$species)

```

Part 2. Tidy the data

Use `mutate()` with `if_else()` to make a new column that categorizes penguins with body mass less than or equal to 4050 as "smaller", otherwise "larger".  
```{r}

penguins_mass_cat <- penguins_df %>% 
  mutate(size_cat = 
    if_else(body_mass_g <= 4050, "smaller", "larger")) 

head(penguins_mass_cat)
```


Use`group_by()` and `summarise()` create a table that will display the counts of species for penguins in the "larger" category (Hint: you will need to also use a `filter()`. 

Then use `mutate()` to add a new column of calculated values from the summarised table to display the proportion of penguins of each species.


```{r final_table}

final_table <- penguins_mass_cat %>% 
  filter(size_cat == "larger") %>% 
  group_by(species) %>% 
  summarise(di_spec_count= n()) %>% 
  # add a mutate function to the summarised table. 
  mutate(prop_of_species = 
           round((di_spec_count / sum(di_spec_count)),2))


```


Bonus: print final table in a print-friendly fashion using `kable()`
```{r kable}
library(knitr)
kable(final_table, "simple",col.names=c('Species','Number of large penguins','Percent of all large penguins'))
```


