---
title: "PHW251 - F21_midterm_prep"
author: "Avery Richards"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
# load tidyverse
library(tidyverse)

# load penguins dataset
install.packages("palmerpenguins")
library(palmerpenguins)

```

Part 1. Import and explore the data

```{r import_summary}


```

Return a summary of the distribution of penguin body mass observations

```{r}


```

Return the unique names of species of penguins observed in the study
```{r}


```

Part 2. Tidy the data

Use `mutate()` with `if_else()` to make a new column that categorizes penguins with body mass less than or equal to 4050 as "smaller", otherwise "larger".  
```{r}


```


Use`group_by()` and `summarise()` create a table that will display the counts of species for penguins in the "larger" category (Hint: you will need to also use a `filter()`. 

Then use `mutate()` to add a new column of calculated values from the summarised table to display the proportion of penguins of each species.


```{r final_table}



```


Bonus: print final table in a print-friendly fashion using `kable()`
```{r kable}
library(knitr)


```


