---
title: 'Live Session #2 - 9/23/2021'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Importing a file
Import a dataset from the California Safe Cosmetics Program (CSCP) in the 
California Department of Public Health. The primary purpose of the CSCP is to 
collect information on hazardous and potentially hazardous ingredients in cosmetic 
products sold in California and to make this information available to the public. 

Our goal is to look at data for products that have been reported to contain 
titanium dioxide since 2015. Specifically, we are interested in identifying 
whether or not titanium dioxide has been removed from products, and if so, how 
long after being reported it was removed. Our end goal is to produce a sentence 
to summarize the # of products that have had the chemical removed.

```{r import}
file_path <- "https://data.chhs.ca.gov/dataset/596b5eed-31de-4fd8-a645-249f3f9b19c4/resource/57da6c9a-41a7-44b0-ab8d-815ff2cd5913/download/cscpopendata.csv"

#import full file just to see what it looks like
cos_full <- read_csv(
  file_path
)

```

Let's import the file again, but this time only import the following columns and ensure they are the corresponding data types:
* CDPHId - character
* BrandName - character
* PrimaryCategory - character 
* ChemicalName - character
* InitialDateReported - date
* ChemicalDateRemoved - character (will convert later)
```{r custom_import}

#import only the columns we need
cos_subset <- read_csv(
  file_path,
  col_types = cols_only(CDPHId = col_character(), ProductName = col_character(), BrandName = col_character(), PrimaryCategory = col_character(),ChemicalName = col_character(), InitialDateReported = col_date(format="%m/%d/%Y"), ChemicalDateRemoved = col_character()),
)

```

## Quick exploration
* Structure of dataset/table
* Unique chemical names
* Frequency of product type
* Distribution of dates reported
```{r}

#structure of dataset
str(cos_subset)

#Unique chemical names - return list of values - commented out to limit output when knitted
# unique(cos_subset$ChemicalName)

#Frequency of product type - return list of values and frequencies
table(cos_subset$PrimaryCategory)

#Distribution of dates reported
summary(cos_subset$InitialDateReported)



```


## Tidying the table

Create a subset with:
* only records for Titanium dioxide -
  + Note that there two different relevant categories: Titanium dioxide AND Titanium dioxide (airborne, unbound particles of respirable size)
* only the following columns: BrandName, ProductName, ChemicalName, InitialDateReported, ChemicalDateRemoved
* only records with initial report date since 2015

Try this using base R, ideally in one line of code:
```{r tidy_base}

#using subset - option for subsetting rows, but does not address the column issue!!
cos_subset2 <- subset(cos_subset, ChemicalName %in% c("Titanium dioxide","Titanium dioxide (airborne, unbound particles of respirable size)") & year(InitialDateReported)>=2015)

#selecting columns
cos_subset3 <- cos_subset[ ,c("BrandName", "ProductName", "ChemicalName", "InitialDateReported", "ChemicalDateRemoved") ]

#doing both in one "line"
cos_subset3b <- cos_subset[which(cos_subset$ChemicalName %in% c("Titanium dioxide","Titanium dioxide (airborne, unbound particles of respirable size)") & year(cos_subset$InitialDateReported)>=2015) , 2:6]

```

Now try using the tidyverse. Once you have created the subset, remove any duplicate lines/rows and sort table to show by most recent date reported.
```{r tidy_dplyr}

cos_tidy <- cos_subset %>%
  select(-c(CDPHId)) %>%
  filter(ChemicalName %in% c("Titanium dioxide","Titanium dioxide (airborne, unbound particles of respirable size)") & year(InitialDateReported)>=2015) %>%
  distinct() %>%
  arrange(desc(InitialDateReported))

```


## Create new columns

* Convert ChemicalDateRemoved from character to date 
* Create a new variable (removed_yn) indicating if the chemical has been removed. If there is no chemical removed date, then we will assume it has not been removed.
* Create a removal category (removal_cat): - skipped in live session
  + Removed before reported
  + Removed within 30 days of reporting
  + Removed within year of reporting
  + Not removed

```{r new_columns}

cos_removed <- cos_tidy %>%
  mutate(
    ChemicalDateRemoved_new = as_date(ChemicalDateRemoved, format="%m/%d/%Y"),
    removed_yn = if_else(is.na(ChemicalDateRemoved_new),"No","Yes"),
    days_to_removal = ChemicalDateRemoved_new - InitialDateReported,
    removal_cat = case_when(
      removed_yn == "Yes" & ChemicalDateRemoved_new<InitialDateReported ~ "Removed before reported",
      removed_yn == "Yes" & days_to_removal<30 ~ "Removed within 30 days",
      removed_yn == "Yes" & days_to_removal<365 ~ "Removed within 365 days",
      removed_yn == "Yes" ~ "Removed more than 365 days",
      TRUE ~ "Not removed"
    )
  )

#check out newly created variable
table(cos_removed$removal_cat)

```

## Final Summary

Return two vectors: 
1. Vector of all unique products with titanium dioxide (try this using base R)
2. Vector of all unique products that removed Titanium dioxide (try this using tidyverse)

```{r create_vectors}

#one way to create a vector using base R
cos_all_products_td <- unique(cos_removed$ProductName)

#create a vector using dplyr
cos_good <- cos_removed %>%
  filter(removed_yn=="Yes") %>%
  select(ProductName) %>%
  distinct() %>%
  pull()


```

Use the vectors above to store the total # products with titanium dioxide AND # products with titanium dioxide removed 
```{r store_totals}

total <- length(cos_all_products_td)

removed <- length(cos_good)

```

Based on the total #/% of products with titanium dioxide removed, create an autogenerated summary based on the proportion of products with titanium dioxide removed. If >10% products, make the summary positive; otherwise, express disappointment.
```{r summary}

if(removed/total*100 > 10){
  
  headline <- "We are making progress! >10% of products have the chemical removed"
  
} else {
  
  headline <- "We still have a lot of work to do :("
  
}

headline

```

```{r ad_hoc_questions}

#Add percent - this version will convert it for a string
paste0(round(removed/total*100,2),"%")

#Check out scales package to keep it numeric


```

